---
title: "WASH & Waterborne Illness — Hararghe, Ethiopia"
author: "Deepanshu Goel"
date: "`r Sys.Date()`"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: flatly
    self_contained: true
    mathjax: null
    navbar:
      - { title: "GitHub Repository", href: "https://github.com/dee-2197/DATA_555_dashboard", align: right, icon: "fa-github" }
---

<style>
/* ── Make takeaways visible & responsive ────────────────────────────────── */
.takeaway-box {
  font-size: 13px;
  line-height: 1.65;
  color: #2C3E50;
  background: #f0f5fb;
  border-left: 5px solid #1B4F8A;
  padding: 12px 16px;
  border-radius: 0 6px 6px 0;
  margin: 0;
}
.takeaway-box b { color: #1B4F8A; }

/* Allow content to overflow into visible space */
.section.level3 { overflow: visible !important; padding-bottom: 6px; }
.chart-wrapper { overflow: visible !important; }
.chart-shim    { overflow: visible !important; }

/* ── Mobile responsive ──────────────────────────────────────────────────── */
@media (max-width: 900px) {
  .takeaway-box { font-size: 12px; padding: 10px 12px; }
  .js-plotly-plot, .plotly { width: 100% !important; }
  .col-sm-6 { width: 100% !important; }
  .row { display: block !important; }
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(flexdashboard)
library(tidyverse)
library(readxl)
library(plotly)

ethiopia <- read_excel("Ethopia_WASH.xlsx")
names(ethiopia)[96] <- "diarrhea_cases"
names(ethiopia)[78] <- "handwashing_practice"
names(ethiopia)[18] <- "education_status"
names(ethiopia)[14] <- "sex"
names(ethiopia)[15] <- "age"

ethiopia <- ethiopia |>
  mutate(
    diarrhea_cases       = trimws(diarrhea_cases),
    handwashing_practice = trimws(handwashing_practice),
    education_status     = trimws(education_status),
    sex                  = trimws(sex),
    age                  = as.numeric(age),
    hw_label = case_when(
      handwashing_practice == "No"                           ~ "No Facility",
      handwashing_practice == "Yes, with water only"         ~ "Water Only",
      handwashing_practice == "Yes, with soap/ash and water" ~ "Soap & Water"
    ),
    hw_label = factor(hw_label, levels = c("No Facility","Water Only","Soap & Water")),
    education_label = case_when(
      str_detect(education_status, "Illiterate")     ~ "Illiterate",
      str_detect(education_status, "read and write") ~ "Read & Write",
      str_detect(education_status, "Some Primary")   ~ "Some Primary",
      str_detect(education_status, "High school")    ~ "High School",
      str_detect(education_status, "College")        ~ "College"
    ),
    education_label = factor(education_label,
      levels = c("Illiterate","Read & Write","Some Primary","High School","College"))
  )

C_NO  <- "#E87722"
C_YES <- "#1B4F8A"

make_chart_data <- function(data, group_var, group_levels) {
  ct <- data |>
    filter(!is.na(.data[[group_var]]), !is.na(diarrhea_cases)) |>
    group_by(across(all_of(c(group_var, "diarrhea_cases")))) |>
    summarise(n = n(), .groups = "drop") |>
    group_by(across(all_of(group_var))) |>
    mutate(total = sum(n), proportion = round(n / total * 100, 1)) |>
    ungroup()

  full <- expand_grid(
    !!group_var := factor(group_levels, levels = group_levels),
    diarrhea_cases = c("No", "Yes")
  )
  full |>
    left_join(ct, by = c(group_var, "diarrhea_cases")) |>
    replace_na(list(n = 0, total = 0, proportion = 0))
}

make_bar_fig <- function(chart_data, group_var, group_levels,
                         x_title, plot_title) {
  fig <- plot_ly()
  for (outcome in c("No", "Yes")) {
    d <- chart_data |> filter(diarrhea_cases == outcome)
    fig <- fig |>
      add_bars(
        x = d[[group_var]], y = ~n, data = d,
        name = paste("Diarrhea:", outcome),
        marker = list(color = if (outcome == "No") C_NO else C_YES,
                      opacity = 0.9,
                      line = list(color = "white", width = 1)),
        text = ~paste0(
          "<b>", d[[group_var]], "</b><br>",
          "Diarrhea: <b>", outcome, "</b><br>",
          "Count: <b>", n, "</b><br>",
          "Group total: ", total, "<br>",
          "Proportion: ", proportion, "%"
        ),
        textposition = "none",
        hoverinfo    = "text"
      )
  }

  fig |>
    layout(
      title  = list(text = paste0("<b>", plot_title, "</b>"),
                    font = list(size = 13, family = "Arial"), x = 0.5),
      xaxis  = list(title = list(text = x_title, font = list(size = 12)),
                    tickfont = list(size = 12), automargin = TRUE,
                    categoryorder = "array", categoryarray = group_levels,
                    tickangle = 0),
      yaxis  = list(type = "log",
                    title = list(text = "Households (Log Scale)",
                                 font = list(size = 12)),
                    tickfont = list(size = 11),
                    gridcolor = "rgba(200,200,200,0.4)",
                    automargin = TRUE,
                    tickvals = c(1,2,5,10,20,50,100,200,500,1000),
                    ticktext = c("1","2","5","10","20","50","100","200","500","1000")),
      barmode = "group", bargap = 0.25, bargroupgap = 0.08,
      legend  = list(title = list(text = "<b>Diarrhea</b>"),
                     orientation = "h", x = 0.5, xanchor = "center", y = -0.30),
      plot_bgcolor  = "white", paper_bgcolor = "white",
      font   = list(family = "Arial"),
      margin = list(t = 60, b = 110, l = 55, r = 20),
      autosize = TRUE,
      updatemenus = list(list(
        type = "buttons", direction = "right",
        x = 0.5, xanchor = "center", y = 1.13, yanchor = "top",
        showactive = TRUE,
        buttons = list(
          list(label = "Both",         method = "update", args = list(list(visible = list(TRUE, TRUE)))),
          list(label = "Diarrhea: No", method = "update", args = list(list(visible = list(TRUE, FALSE)))),
          list(label = "Diarrhea: Yes",method = "update", args = list(list(visible = list(FALSE, TRUE))))
        )
      )),
      annotations = list(
        list(text = "Filter:", x = 0.22, xref = "paper",
             y = 1.135, yref = "paper", showarrow = FALSE,
             font = list(size = 11))
      )
    ) |>
    config(displayModeBar = TRUE,
           modeBarButtonsToRemove = c("zoom2d","pan2d","select2d","lasso2d",
                                      "zoomIn2d","zoomOut2d","autoScale2d"),
           responsive = TRUE)
}
```

Dashboard {data-icon="fa-chart-bar"}
=====================================================================

Row {data-height=420}
-------------------------------------

### Figure 1: Handwashing Practice vs. Diarrhea Cases

```{r fig1}
hw_data <- make_chart_data(ethiopia, "hw_label", levels(ethiopia$hw_label))
make_bar_fig(hw_data, "hw_label", levels(ethiopia$hw_label),
             "Handwashing Practice",
             "Does Handwashing Practice Affect Diarrhea Risk?")
```

### Figure 2: Sex vs. Diarrhea Cases

```{r fig2}
sex_data <- make_chart_data(ethiopia, "sex", c("Female", "Male"))
make_bar_fig(sex_data, "sex", c("Female", "Male"),
             "Sex",
             "Does Sex Affect Diarrhea Risk?")
```

Row {data-height=110}
-------------------------------------

###

<div class="takeaway-box">
<b>Figure 1 &mdash; Key Takeaways:</b> Households with <b>no handwashing facility</b> had the highest diarrhea prevalence (17 cases, 3.0%), compared to water-only (6 cases, 2.8%). <b>Zero diarrhea cases</b> were observed in soap &amp; water households &mdash; the strongest protective WASH behavior. Handwashing with soap appears to be a <b>key modifiable risk factor</b> for waterborne illness.
</div>

###

<div class="takeaway-box">
<b>Figure 2 &mdash; Key Takeaways:</b> The sample was predominantly <b>female</b> (n=856, 87.6%), reflecting the household survey design. Diarrhea proportions were similar across sexes: <b>2.5% (female) vs. 1.7% (male)</b> &mdash; sex alone is not a strong predictor of diarrhea risk. The log scale is especially useful given the large group size imbalance (856 vs. 121).
</div>

Row {data-height=420}
-------------------------------------

### Figure 3: Education Status vs. Diarrhea Cases

```{r fig3}
edu_data <- make_chart_data(ethiopia, "education_label", levels(ethiopia$education_label))
make_bar_fig(edu_data, "education_label", levels(ethiopia$education_label),
             "Educational Status",
             "Does Education Status Affect Diarrhea Risk?")
```

### Figure 4: Age Distribution by Diarrhea Status (years)

```{r fig4}
box_colors <- c("Yes" = "#1B4F8A", "No" = "#E87722")

fig4 <- plot_ly()
for (outcome in c("Yes", "No")) {
  d <- ethiopia |> filter(diarrhea_cases == outcome, !is.na(age))
  five_num <- quantile(d$age, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
  fig4 <- fig4 |>
    add_boxplot(
      y = ~age, data = d,
      name      = paste("Diarrhea:", outcome),
      x         = outcome,
      marker    = list(color = box_colors[[outcome]]),
      line      = list(color = box_colors[[outcome]]),
      fillcolor = box_colors[[outcome]],
      opacity   = 0.80,
      text = paste0(
        "<b>Diarrhea: ", outcome, "</b><br>",
        "Min: <b>",    five_num[1], " yrs</b><br>",
        "Q1: <b>",     five_num[2], " yrs</b><br>",
        "Median: <b>", five_num[3], " yrs</b><br>",
        "Q3: <b>",     five_num[4], " yrs</b><br>",
        "Max: <b>",    five_num[5], " yrs</b>"
      ),
      hoverinfo = "text"
    )
}

fig4 |>
  layout(
    title  = list(text = "<b>Age Distribution by Diarrhea Status</b>",
                  font = list(size = 13, family = "Arial"), x = 0.5),
    xaxis  = list(title = list(text = "Diarrhea Status", font = list(size = 12)),
                  tickfont = list(size = 11), automargin = TRUE,
                  categoryorder = "array", categoryarray = c("Yes", "No")),
    yaxis  = list(title = list(text = "Age (years)", font = list(size = 11)),
                  tickfont = list(size = 10),
                  gridcolor = "rgba(200,200,200,0.4)",
                  automargin = TRUE, zeroline = FALSE),
    showlegend    = FALSE,
    plot_bgcolor  = "white", paper_bgcolor = "white",
    font   = list(family = "Arial"),
    margin = list(t = 50, b = 60, l = 55, r = 20),
    autosize = TRUE
  ) |>
  config(displayModeBar = TRUE,
         modeBarButtonsToRemove = c("zoom2d","pan2d","select2d","lasso2d",
                                    "zoomIn2d","zoomOut2d","autoScale2d"),
         responsive = TRUE)
```

Row {data-height=110}
-------------------------------------

###

<div class="takeaway-box">
<b>Figure 3 &mdash; Key Takeaways:</b> Diarrhea cases were concentrated in the <b>illiterate</b> (n=19, 2.5%) and <b>some primary</b> (n=3, 5.0%) groups. <b>Zero cases</b> in high school/preparatory and college/university groups suggest a strong protective educational gradient. Education may be an important <b>effect modifier</b> of the WASH&ndash;diarrhea relationship.
</div>

###

<div class="takeaway-box">
<b>Figure 4 &mdash; Key Takeaways:</b> Median age was <b>30 years</b> (IQR: 25&ndash;39.5) in diarrhea cases vs. <b>35 years</b> (IQR: 28&ndash;40) in non-cases. The overlapping IQRs suggest <b>age alone is not a strong predictor</b> of diarrhea risk. Hover over each box to explore the full 5-number summary.
</div>

Row {data-height=70}
-------------------------------------

###

<p style="text-align:center; font-size:13px; color:#444; padding: 6px 16px; line-height:1.55; margin: 0;">
<b>Why this matters:</b> Diarrheal disease remains a leading cause of child morbidity in rural Ethiopia. Identifying modifiable WASH risk factors &mdash; particularly handwashing with soap &mdash; provides actionable targets for community health interventions and policy.
&nbsp;|&nbsp;
<a href="https://github.com/dee-2197/DATA_555_dashboard" target="_blank" style="color:#1B4F8A; font-weight:600;">📎 Source Code on GitHub</a>
</p>

About the Data {data-icon="fa-info-circle"}
=====================================================================

Row
-------------------------------------

### Dataset Description

**Data Source:** Household survey dataset from Hararghe, Ethiopia (confidential — not publicly available).

**Sample Size:** 977 households.

**Data Collection:** Data were collected via structured household interviews using a cross-sectional survey design targeting household-level WASH practices and health outcomes in the Hararghe region of Ethiopia.

**Study Population:** Household respondents (87.6% female) aged 8–78 years residing in Hararghe, Ethiopia; 76.5% of respondents were illiterate.

**Outcome:** Diarrhea cases reported in the past month (Yes: 23 households, 2.4%; No: 954 households, 97.6%).

**Key Exposures:** Handwashing practice (no facility / water only / soap & water), sex, age, and educational status.

---

**Why this matters:** Diarrheal disease remains a leading cause of child morbidity in rural Ethiopia. Identifying modifiable WASH risk factors — particularly handwashing with soap — provides actionable targets for community health interventions and policy makers working to reduce waterborne illness burden in the region.

---

**Source Code:** The complete source code for this dashboard is available at [github.com/dee-2197/DATA_555_dashboard](https://github.com/dee-2197/DATA_555_dashboard).
